---
title: "Single sample analysis"
author:
  - Pratibha Panwar
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc_float: true
  BiocStyle::pdf_document: default
package: clustSIGNAL
vignette: |
  %\VignetteIndexEntry{1. Single sample analysis}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>"
)
```

# SeqFISH mouse embryo data analysis

```{r}
# load required packages
library(clustSIGNAL)
library(distances)
library(cluster)
library(aricode)
library(dplyr)
library(ggplot2)
library(patchwork)
```

```{r}
data(mouseEmbryo2)
spe
```

```{r}
names(colData(spe))
```

To run clustSIGNAL, we need the column names of sample and cell labels in the colData dataframe of the spatial experiment object. Here, the cell labels are in the column 'uniqueID' and sample labels are in 'embryo' column. 

# Running clustSIGNAL

```{r}
set.seed(100)
samples = "embryo"
cells = "uniqueID"
res_emb = clustSIGNAL(spe, samples, cells, outputs = "a")
```

The output variable is a list that can contain dataframe of cluster names, matrix of NN neighbours of each cell, final spe object, or a combination of these, depending on the choice of 'outputs' selected.

```{r}
names(res_emb)
```

```{r}
head(res_emb$clusters, n = 3)
```

```{r}
spe = res_emb$spe_final
spe
```

# Calculating clustering metrics

```{r}
# calculating silhouette width
clusts <- as.numeric(as.character(spe$reCluster))
cXg_mat <- t(as.matrix(logcounts(spe)))
distMat <- distances(cXg_mat)
silCluster <- as.matrix(silhouette(clusts, distMat))
spe$rcSil <- silCluster[, 3]

# for datasets with annotated cell type information, we can also calculate 
# metrics like adjusted rand index (ARI) and normalised mutual information (NMI)
as.data.frame(colData(spe)) %>%
  summarise(ARI = aricode::ARI(celltype_mapped_refined, reCluster),
            NMI = aricode::NMI(celltype_mapped_refined, reCluster),
            ASW = mean(rcSil),
            min_Entropy = min(entropy),
            max_Entropy = max(entropy),
            mean_Entropy = mean(entropy))
```

# Visualising clustSIGNAL outputs

```{r}
colors = c("#635547", "#8EC792", "#9e6762", "#FACB12", "#3F84AA", "#0F4A9C", 
           "#ff891c", "#EF5A9D", "#C594BF", "#DFCDE4", "#139992", "#65A83E", 
           "#8DB5CE", "#005579", "#C9EBFB", "#B51D8D", "#532C8A", "#8870ad", 
           "#cc7818", "#FBBE92", "#EF4E22", "#f9decf", "#c9a997", "#C72228", 
           "#f79083", "#F397C0", "#DABE99", "#c19f70", "#354E23", "#C3C388",
           "#647a4f", "#CDE088", "#f7f79e", "#F6BFCB", "#7F6874", "#989898", 
           "#1A1A1A", "#FFFFFF", "#e6e6e6", "#77441B", "#F90026", "#A10037", 
           "#DA5921", "#E1C239", "#9DD84A")
```

## Entropy spread and distribution

```{r}
# Histogram of entropy spread
hst_ent <- as.data.frame(colData(spe)) %>%
  ggplot(aes(entropy)) +
  geom_histogram(binwidth = 0.05) +
  ggtitle("A") +
  labs(x = "Entropy", y = "Number of regions") +
  theme_grey() +
  theme(text = element_text(size = 12))

# Spatial plot showing sample entropy distribution
spt_ent <- as.data.frame(colData(spe)) %>%
  ggplot(aes(x = spatialCoords(spe)[, 1], 
             y = -spatialCoords(spe)[, 2])) +
  geom_point(size = 0.5, 
             aes(colour = entropy)) +
  scale_colour_gradient2("Entropy", low = "grey", high = "blue") +
  scale_size_continuous(range = c(0, max(spe$entropy))) +
  ggtitle("B") +
  labs(x = "x-coordinate", y = "y-coordinate") +
  theme_classic() +
  theme(text = element_text(size = 12))

hst_ent + spt_ent
```

The spread (A) and spatial distribution (B) of region entropy measures can be very useful in assessing the tissue composition of samples - low entropy regions are more homogeneous with domain-like structure, whereas high entropy regions are heterogeneous with more uniform distribution of cells.

## clustSIGNAL clusters visualisation

```{r}
df_ent = as.data.frame(colData(spe))

# spatial plot
spt_clust <- df_ent %>%
  ggplot(aes(x = spatialCoords(spe)[, 1], 
             y = -spatialCoords(spe)[, 2])) +
  geom_point(size = 0.5, aes(colour = reCluster)) +
  scale_color_manual(values = colors) +
  ggtitle("A") +
  labs(x = "x-coordinate", y = "y-coordinate") +
  guides(color = guide_legend(title = "Clusters", 
                              override.aes = list(size = 3))) +
  theme_classic() +
  theme(text = element_text(size = 12))

# calculating median entropy of each cluster
celltype_ent = df_ent %>%
  group_by(as.character(reCluster)) %>%
  summarise(meanEntropy = median(entropy))
# reordering clusters by their median entropy
# low to high median entropy
cellOrder = celltype_ent$meanEntropy
names(cellOrder) = celltype_ent$`as.character(reCluster)`
cellOrder = sort(cellOrder)
df_ent$reCluster = factor(df_ent$reCluster, levels = names(cellOrder))
# box plot of cluster entropy
colors_ent = colors[as.numeric(names(cellOrder))]
box_clust <- df_ent %>%
  ggplot(aes(x = reCluster, y = entropy, fill = reCluster)) +
  geom_boxplot() +
  scale_fill_manual(values = colors_ent) +
  ggtitle("B") +
  labs(x = "clustSIGNAL clusters", y = "Entropy") +
  theme_classic() +
  theme(legend.position = "none",
        text = element_text(size = 12),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

spt_clust + box_clust + patchwork::plot_layout(guides = "collect", widths = c(1, 2))
```

The spatial location (A) and entropy distribution (B) of the clusters provide spatial context of the cluster cells and their neighbourhoods, as well as the compositions of the neighbouhoods.
