---
title: "Multisample analysis"
author:
  - Pratibha Panwar
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc_float: true
  BiocStyle::pdf_document: default
package: clustSIGNAL
vignette: |
  %\VignetteIndexEntry{2. Multisample analysis}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>"
)
```

# MERFISH mouse preoptic hypothalamus data analysis

```{r}
# load required packages
library(clustSIGNAL)
library(distances)
library(cluster)
library(aricode)
library(dplyr)
library(ggplot2)
library(patchwork)
```

```{r}
data(mousePH_subset)
spe2
```

```{r}
names(colData(spe2))
```

To run clustSIGNAL, we need the column names of sample and cell IDs in the colData dataframe of the spatial experiment object. Here, the cell IDs are in the column 'Cell_ID' and sample IDs are in 'samples' column.

# Running clustSIGNAL

```{r}
set.seed(101)
samples = "samples"
cells = "Cell_ID"
res_hyp = clustSIGNAL(spe2, samples, cells, outputs = "a")
```

```{r}
spe2 = res_hyp$spe_final
spe2
```

# Calculating clustering metrics

```{r}
samplesList <- levels(spe2[[samples]])
```

```{r}
# calculating silhouette width per sample
silWidthRC <- matrix(nrow = 0, ncol = 3)
for (s in samplesList) {
  speX <- spe2[, spe2[[samples]] == s]
  clust_sub <- as.numeric(as.character(speX$reCluster))
  cXg <- t(as.matrix(logcounts(speX)))
  distMat <- distances(cXg)
  silCluster <- as.matrix(silhouette(clust_sub, distMat))
  silWidthRC <- rbind(silWidthRC, silCluster)
}
spe2$rcSil <- silWidthRC[, 3]
```

```{r}
# for datasets with annotated cell type information, we can also calculate 
# metrics like adjusted rand index (ARI) and normalised mutual information (NMI)
as.data.frame(colData(spe2)) %>%
  group_by(samples) %>%
  summarise(ARI = aricode::ARI(Cell_class, reCluster),
            NMI = aricode::NMI(Cell_class, reCluster),
            ASW = mean(rcSil),
            min_Entropy = min(entropy),
            max_Entropy = max(entropy),
            mean_Entropy = mean(entropy))
```

# Visualising clustSIGNAL outputs

```{r}
colors = c("#635547", "#8EC792", "#9e6762", "#FACB12", "#3F84AA", "#0F4A9C", 
           "#ff891c", "#EF5A9D", "#C594BF", "#DFCDE4", "#139992", "#65A83E", 
           "#8DB5CE", "#005579", "#C9EBFB", "#B51D8D", "#532C8A", "#8870ad", 
           "#cc7818", "#FBBE92", "#EF4E22", "#f9decf", "#c9a997", "#C72228", 
           "#f79083", "#F397C0", "#DABE99", "#c19f70", "#354E23", "#C3C388",
           "#647a4f", "#CDE088", "#f7f79e", "#F6BFCB", "#7F6874", "#989898", 
           "#1A1A1A", "#FFFFFF", "#e6e6e6", "#77441B", "#F90026", "#A10037", 
           "#DA5921", "#E1C239", "#9DD84A")
```

## Entropy spread and distribution

```{r}
# Histogram of entropy spread
hst_ent <- as.data.frame(colData(spe2)) %>%
  ggplot(aes(entropy)) +
  geom_histogram(binwidth = 0.05) +
  ggtitle("A") +
  facet_wrap(vars(samples), nrow = 1) +
  labs(x = "Entropy", y = "Number of regions") +
  theme_grey() +
  theme(text = element_text(size = 12))

# Spatial plot showing sample entropy distribution
spt_ent <- as.data.frame(colData(spe2)) %>%
  ggplot(aes(x = spatialCoords(spe2)[, 1], 
             y = -spatialCoords(spe2)[, 2])) +
  geom_point(size = 0.5, 
             aes(colour = entropy)) +
  scale_colour_gradient2("Entropy", low = "grey", high = "blue") +
  scale_size_continuous(range = c(0, max(spe2$entropy))) +
  facet_wrap(vars(samples), scales = "free", nrow = 1) +
  ggtitle("B") +
  labs(x = "x-coordinate", y = "y-coordinate") +
  theme_classic() +
  theme(text = element_text(size = 12),
        axis.text.x = element_text(angle = 90, vjust = 0.5))

hst_ent / spt_ent
```

In multisample analysis, the spread (A) and spatial distribution (B) of region entropy measures can be useful in assessing and comparing the tissue structure in the samples.

## clustSIGNAL clusters visualisation

```{r}
df_ent = as.data.frame(colData(spe2))

# spatial plot
spt_clust <- df_ent %>%
  ggplot(aes(x = spatialCoords(spe2)[, 1], 
             y = -spatialCoords(spe2)[, 2])) +
  geom_point(size = 0.5, aes(colour = reCluster)) +
  scale_color_manual(values = colors) +
  facet_wrap(vars(samples), scales = "free", nrow = 1) +
  labs(x = "x-coordinate", y = "y-coordinate") +
  guides(color = guide_legend(title = "Clusters", 
                              override.aes = list(size = 3))) +
  theme_classic() +
  theme(text = element_text(size = 12),
        axis.text.x = element_text(angle = 90, vjust = 0.5))

box_clust = list()
for (s in samplesList) {
  df_ent_sub = as.data.frame(colData(spe2)[spe2[[samples]] == s, ])
  # calculating median entropy of each cluster in a sample
  celltype_ent = df_ent_sub %>%
    group_by(as.character(reCluster)) %>%
    summarise(meanEntropy = median(entropy))
  # reordering clusters by their median entropy
  # low to high median entropy
  cellOrder = celltype_ent$meanEntropy
  names(cellOrder) = celltype_ent$`as.character(reCluster)`
  cellOrder = sort(cellOrder)
  df_ent_sub$reCluster = factor(df_ent_sub$reCluster, levels = names(cellOrder))
  
  # box plot of cluster entropy
  colors_ent = colors[as.numeric(names(cellOrder))]
  box_clust[[s]] <- df_ent_sub %>%
    ggplot(aes(x = reCluster, y = entropy, fill = reCluster)) +
    geom_boxplot() +
    scale_fill_manual(values = colors_ent) +
    facet_wrap(vars(samples), nrow = 1) +
    labs(x = "clustSIGNAL clusters", y = "Entropy") +
    ylim(0, NA) +
    theme_classic() +
    theme(legend.position = "none",
          text = element_text(size = 12),
          axis.text.x = element_text(angle = 90, vjust = 0.5))
}

spt_clust / (patchwork::wrap_plots(box_clust[1:3], nrow = 1) + 
               plot_layout(axes = "collect")) + 
  plot_layout(guides = "collect") +
  plot_annotation(title = "Spatial and entropy distributions of clusters")
```

The spatial location (top) and entropy distribution (bottom) of the clusters can be compared in a multisample analysis, providing spatial context of the cluster cells and their neighbourhood compositions in the different samples.
