---
title: "Getting started: clustSIGNAL"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started: clustSIGNAL}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>"
)
```

# Introduction

clustSIGNAL (clustering using Spatially Informed Gene expression with Neighborhood Adapted Learning) is an R package for a spatial clustering method that takes into account the heterogeneity of cell neighborhoods.

```{r}
# load required packages
library(clustSIGNAL)

```

# Using clustSIGNAL with seqFISH mouse embryo data

## run clustSIGNAL

```{r}
set.seed(100)

```

## calculate clustering metrics

```{r}

```

## visualisation

```{r}

```

# Using clustSIGNAL with MERFISH mouse preoptic hypothalamus data

## run clustSIGNAL

```{r}
set.seed(100)

```

## calculate clustering metrics

```{r}

```

## visualisation

```{r}

```

